RSAT

### 2015-05-14 MScCS.tex
COMMIT! innnan ungern

\title{Formatting a Master's Thesis}
\subtitle{A {\LaTeX} class}
%\company{The Corporation AB LTD Inc}
\supervisors{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}{Don Jeer, \href{mailto:djeer@xy.lth.se}{\texttt{djeer@xy.lth.se}}}
%\supervisor{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}
\examiner{Jane Doe, \href{mailto:jane.doe@cs.lth.se}{\texttt{jane.doe@cs.lth.se}}}

@@

\theabstract{
This document describes the Master's Thesis format for the theses carried out at 
the Department of Computer Science, Lund University.

Your abstract should capture, in English, the whole thesis with focus on the problem and solution in 150 words. It should be placed on a separate right-hand page, with an additional \textit{1cm} margin on both left and right. Avoid acronyms, footnotes, and references in the abstract if possible.

Leave a \textit{2cm} vertical space after the abstract and provide a few keywords relevant for your report. Use five to six words, of which at most two should be from the title.
}

\keywords{MSc, template, report, style, structure}

@@

\chapter[Introduction]{Introduction}
%in need of a kidney transplant can not undergo surgery

Many people suffering from renal failure are in need of a kidney transplant. A problem in the health care is that patients can not undergo surgery if they have too much belly fat. It is very hard for this group of people to loose weight, as further described in section \ref{sec:kiney}. Regular exercises and physical activities are therefore something crucial for this group of people. A project group at the Skåne University Hospital has recently been established to help this group of people to loose weight. A question they asked themselves was whether it was possible to use any activity tracking devices in this project.

Sony has developed wearable devices that can track a user’s daily activities. The data collected by these devices are saved online in their database and accessible through an API called the LifeLog API. Sony wants to investigate what different use cases that exists for their wearable devices and has therefore chosen to sponsor this project with smart wristbands for the patients to wear.

itACiH is a research project that is investigating the possibilities of advanced health care at home. As a part of this project a planning-system was developed that currently is in use at the hospital. This planning system is built as a web application. The system is built such that it remotely can fetch data from sensors that are installed at the home of a patient. Because itACiH is already in use at hospital the system developed in this thesis will use the same platform as itACiH for easy later integration at the hospital.

This thesis presents a system that aims to help over-weight patients with renal failure to loose weight. The system is called the Region Skåne Activity tracking, or the RSAT for short. The project group will be called the RSAT-group in this thesis.

@@

A technical evaluation of SONY Lifelog API for use in heath care.

How good is SONY Lifelog  for use in heath care from a technical standpoint?

How good is RSAT?

@@

\end{comment}
\section{Method}%TODO

The method used for the development of the system was by first talking to the RSAT-grope and listen to there wishes. Then making a prof of concept solution to get familiar with the API. By using the experience of from the development of the prof of concept requirements was formed together with the RSAT-group. The approach for the development was to use iterations. Each iteration resulted in a demo for the RSAT-group. During the demo feedback was given from the RSAT-group. The experience from the development of the system lay down the ground for the evaluation of both Lifelog and RSAT.

%As the purpose of this thesis is twofold; (1) to investigate in what ways the Sony Lifelog API is useful in the health care and, (2) to build an actual system based on the API, the work carried out in this thesis is divided into three main parts, one research part, one development part and one evaluation part.

% This thesis is divided into three main parts; one research part, one development part and one evaluation part. In the research part facts about other systems similar to the one to be built will be gathered. Information about other activity trackers will also be collected in order to evaluate the Sony Lifelog and to compare it with other platforms. In the development, the development of the system will be performed and explained. In the evaluation part the RSAT will be evaluated and compared to other existing systems. The Sony Lifelog API will also be evaluated and compared to other API:s.

% During the project, three different prototypes will be built and the development cycle of each prototype will consist of the following steps:

\begin{enumerate}
\item Identify the requirements for the current prototype
\item Implement functionality in the system such that the system fulfills the requirements.
\item Test the developed prototype and publish it such that others can test our software.
\item Receive feedback from testers and use it as input to the next prototype. 
\item Receive new requirements on the system, if the requirements are feasible, add them to a later iteration. 
\end{enumerate}

@@

The fist part of this thesis gives an overview of the landscape that RSAT acts in. Fist a medical background is given. Then a technical background describing the techniques used in RSAT. After that works related to the thesis is described. First different warble technologies and platforms for aggregation of activity data and  systems similar to RSAT.

The second part contains a description of the development of RSAT. From the initial thoughts in the RSAT group to the last prototype. Telling about all decision and struggles.

@@

%Third part evaluate both Lifelog and RSAT by comparing them to solutions and standards described in the first part of the thesis.

%In the first part, an overview of the current state of the art of wearable technology is given. Also different platforms that can be used to aggregate health data are presented. In the second part, an actual system is developed. At last the system is evaluated and compared to other systems. %TODO

@@

The kidneys have the responsibility of filtering waste products from the blood in the body\cite{NKDEPBASICS}. When the kidney function is too low to filter the blood, dialysis and in the end a kidney transplant, is needed.

In this chapter an introduction to chronic kidney disease is given. The following chapter begins with a short introduction to chronic conditions in general. Chapter 2.2 discusses chronic kidney disease especially.

@@

%Treating chronic condition is a huge part of the health care, especially in the western world.

% todo: kan man skriva "go to treating" är go to kanske lite "sv-engelskt"? Tror klar hittar du fel jens fixa.

The costs in the health care are ever increasing. A huge part of the health care and the economic costs associated with it addresses patients with chronic conditions. More than 60\% of all deaths in the world is due to this types of complications. In Sweden 80\% of all the health care costs go to treating these conditions\cite{SmartCare}. A study shows that three types of lifestyle behaviors leads to four chronic diseases. These diseases accounts for about 50\% of all the deaths worldwide, see figure \ref{fig:chronic-fig}\cite{callaway2015quantified}.

@@

Chronic kidney disease (CKD) is a gradual loss of kidney function over time\cite{KIDNEYFOUNDATION}. There are no specific symptoms of the disease but illness, tiredness and a reduced appetite are all common symptoms. As the kidney function is worsening, waste products builds to high levels in the blood that may cause complications such as high blood pressure,  muscle- and bone mass loss. The risk of getting other diseases such as cardiovascular disease, ericarditis and anemia is also growing.  When the waste products in a patient’s body becomes so high that the patient starts to get sick from it, the patient requires dialysis\cite{KIDNEYFOUNDATION}.

@@

Despite that CKD is considered to be an irreversible condition, the progress of the disease can be slowed down. The National Kidney Disease Education Program (NKDEP) lists three strategies that can be used to slow down the progress\cite{NKDEPPROG}. One of those strategies is lifestyle interventions and addresses different health-promoting behaviors. One of the health-promoting behaviors they discuss is physical activity. NKDEP states that people suffering from CKD has the same recommendations for physical activity as the general population.

Eventually, a patient with CKD must undergo a kidney transplant. The patient must be in good physical shape, as the risk of infections and other diseases otherwise are too high. It is also essential that the patient does not have to much belly fat. Not only does the fat induce a greater risk for the patient's life, it also makes it harder for surgeons to practically perform the transplant. Vad kostar det?!
%http://www.kidney.ca/document.doc?id=4083
%http://health.costhelper.com/kidney-cancer.html

@@

This chapter describes the techniques used to develop RSAT. First Sony Lifelog and techniques used by Lifelog described.The second part of the describes Palcom the platform used by the itACiHi and RSAT.

@@

% Write about the smart armband
The Sony Lifelog application is an Android application which is built around the idea about logging a users lifestyle and daily activities\cite{LifeLogDescr}. The application collects various data about its users through different sensors in a smartphone and SmartWear-devices (Sony SmartWatch or a Sony Smartband) and displays it to the user. Example of data it logs are information about what applications in the smartphone that have been used, what locations that have been visited and information about different physical activities performed during the day.

@@

The Sony SmartBand SWR10 is the chosen wristband for this project. The Sony SWR10 has a movement sensor which recognizes different movements a user is doing\cite{SWR10}. This data is analyzed by Sony and categorized into different activites which are uploaded to a central repository that stores the data. The SmartBand must be connected to an Android device with Android version 4.4 or higher to be able to collect data.

\subsection{The Sony Lifelog API} 
All the data collected in the Sony Lifelog application are automatically uploaded to a central repository that stores the data. This data is globally accessible through a REST-based web-serviceAPI\cite{LifeLogDescr}.

The API has three different endpoints; a profile endpoint, an activities endpoint and a location endpoint. The profile endpoint contains basic information about a user's name, age, height and weight and is entered by the user into the Lifelog application. The activities endpoint contains the different data Lifelog has logged about performed activities throughout the day such as physical activity. All activities has a type and some also has a subtype.

%Examples of data a user can log is data about her sleep or how many steps she has walked during the day. 
The location endpoint provides a list of coordinates specifying what locations a user has been visiting during a time period. Activities and locations in the API are associated with times. A response from the API always includes the time stamp of the earliest and latest included entry. The format of the timestamp follows the ISO 8601 standard.

The API can be accessed by third parties to retrieve collected data about a user. Every request made to the API must be authorized and https must be the chosen communication protocol. The lifelog Web API uses OAuth 2.0 as its authorization mechanism. To connect to the API, a user must login and give explicit permission for the application or the service to access to the data.

The response to a valid request made to the API contains a result-section. This result-section is a set of data represented as json objects. If the requested data is to large to fit in the size of one response message, the data is split up into several messages and a link-field is inserted into the response. This link-field contains a link pointing to the location from where the rest of the data can be fetched.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}
The response can also contain an error message:
403 AccessDenied				you do not have permission to access
401 InsufficientAuthentication	No access token or an invalid access token was sent in the request
404 ResourceNotFound			The data does not exist
422 UnprocessableEntity 		Request could not be parsed or contained invalid values
\end{comment}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The Sony Lifelog API is a REST API. REST is an architectural style for designing API:s and are heavily used in modern web services. The concept is to represent resources as the analogy for the data endpoints and to have a small set of actions interacting with these resources. The REST architectural style has the following principles\cite{pautasso2014restful}:

\begin{itemize}
    \item \emph{Addressabilty} All resources must be addressable with a unique identification. On the web each resource uses a unique URI such that it becomes globally addressable.
    \item \emph{Uniform Interface} The interface to the resources should according to REST be uniformed and simple. On the web the HTTP protocol is used and HTTP methods is used to interact with resources. To read a resources the GET method is used, to add a resource the POST method is used, to modify a resource the PUT method is used and to delete a resource the DELETE method is used.
    \item \emph{Stateless Interaction} This means that there is no session with multiple interactions. Instead every interaction is self-contained. No state is shared between a service and client after an interaction.
    \item \emph{Self-Describing Messages} The requests and responses should contain both the data and the meta data about the format of the data. XML or JSON is usually used to represent data on the web. The Sony Lifelog API uses JSON to represent its data.

\item \emph{Hypermedia} Resources that contain links to other resources use hypermedia. A link can for example a link to the next day in a calender API in the response of the current day use hypermedia.

@@

OAuth 2.0 is an authorization protocol developed by the Internet Engineering Task Force (IETF). It describes an open standard to authorization and enables third-party applications to obtain limited access to a web-service on the behalf of a resource owner\cite{oauth2spec}. To describe the protocol, several key actors needs to be described. These actors are the client, the resource owner, the authorization server and the resource server. IETF gives the following definitions to the roles:

@@

The following example gives a general example of the protocol flow when a client is accessing protected data belonging to a user.

\begin{enumerate}
	\item A client requests authentication from a resource owner. 
	\item The client receives an authorization grant credential from the resource owner.    
	\item The client requests an access token from the authorization server. The server verifies the authorization grant credential and returns an access token. 
	\item The client receives an access token which it can use to access the protected data in the resource server. 
\end{enumerate}

The Authorization Grant is a credential representing the resource owner's authorization. The IETF's OAuth 2.0 specification defines four different grant types; authorization code, implicit, resource owner password credentials and client credentials. The Sony Lifelog API uses the authorization code as its grant type and the other types will therefore not be further discussed\cite{LifeLogAuth}. When the authorization code grant type is used, the resource owner is redirected to an authentication server when the client requests authentication. When the resource owner has granted the request, the resource owner is redirected back to the client together with the authorization code which is appended in the URL. The authentication code is short lived and single-use, meaning that if the code is used more than once, the access tokens related to the code shall be revoked.

The access token is the only credential needed to access a user's protected resource and can be seen as an abstraction layer, since other credentials such as user-name and password must not be provided. An access token is valid until it expires or it is revoked.

A refresh token can be issued together with an access token and is used to refresh expired or revoked access tokens. To refresh an access token, the refresh token is sent to the authorization server, which issues a new access token. A refresh token is meant to be valid for a longer time than an access token but that is up to the server implementing the OAuth-specification to decide. This means that a refresh token in theory could expire before an access token.

\begin{comment}
\section{PalCom}
PalCom is a platform that makes it possible to build software systems which consists of small sub-systems\cite{ist-PalCom}. These sub-systems are called \emph{PalCom Services} and are self-contained, they do not depend on the context or state of other services in the system. Different services are composed together with help of \emph{PalCom Assemblies}. PalCom Assemblies describes how different services operates and communicates in the system. At a conceptual level an assembly can be described as a wire which connects different services. This conceptual wire works the same way as for instance a hdmi-cable between a computer and a projector. 
What an assembly actually does is defining rules for how commands shall be routed between services.

%What an assembly actually does, is that it tells how different services can be reached from different devices and how different services can use each others functionality.

%An Assembly define rules how command shall be roted between services.

A PalCom Service is the unit which performs the actual computation in a PalCom System. Each service has an interface signature describing what commands the service can send and receive, their parameters and the type of the parameters. This signature is broadcasted on the network the service is connected to and is used when assemblies are created to compose different services.

Suppose you were about to build a very primitive chat application in PalCom with the purpose of letting two clients send text messages to each other. Such an application would consists of two identical services. One for each client which in turn would have implemented two different PalCom commands; a send command and a receive command. Both these functions would be described in the interface signature of the chat service. To connect the two services and make the communication between the possible, assemblies would be used. The assembly would state that each servers out command should be connected to each servers in command. Each service has a mechanism which makes it possible to discover other services and to advertise its existence. This mechanism is called the \emph{PalCom Runtime Component}. This would make it possible for the two services to discover each other.

Different services are run on an application called theThing. It has a JVM for running services and an interpreter for assemblies. It also handles all the network configuration and the connection to the PalCom network.

\begin{comment}
\section{Palcom}
Palcom is middleware used to connect systems. Here we define the terms used in this thesis regarding Palcom:

\begin{itemize}
\item \emph{Palcom Service} is a type of self-contained systems. The services contain logic for both computation and interaction with the outside world. Implementation of services is done i Java and can therefore leverage tho whole Java ecosystem.

\item \emph{Palcom Command} is what Palcom Services use to communicate with each other. A service define a number och commands, they are ether \emph{in commands} or \emph{out commands}. \emph{In commands} define the format for command sent to the service. And \emph{out commands} define in the same way the format for command sent from the service. The commands consist of a number och parameter with a name and a data type. An example of a type is "text".

\item \emph{Palcom Assemlies} is rules for how Palcom commands shall be routed in the system. Instead of the service deciding what other service should receive the command the service sends. The assemblies describe this with easy rules like "when service s1 sends command cmdOut send it to CmdOut on service s2". 
\end{itemize}
\end{comment}

\section{Palcom}
The planning system in the itACiH project use Palcom as there platform. PalCom is a platform that makes it possible to build software systems which consists of small sub-systems\cite{ist-PalCom}. These sub-systems are called called \emph{PalCom Services}. Different services are composed together with the help of assemblies. An explanation of what the different component in the PalCom platform are is given below:

\begin{itemize}
\item \emph{PalCom Service} - A PalCom service is a type of self-contained system. The service contains logic for both computations and interactions with the outside world. The implementation of services is made in Java and can therefore leverage the whole Java ecosystem. Every version service has unique id called \emph{serviceId} and every instance of a service has an unique id called \emph{instance}.

\item \emph{Palcom Command} - A PalCom command is what a Palcom Service uses to communicate with other services. A service defines a number of commands, they are either \emph{in-commands} or \emph{out-commands}. An in-command defines the format of a command sent in to the service and an out command defines the format of a command sent from the service. A command consists of zero or more parameters. Each parameter has a name and a data type. The data type defines what type of data that is sent in the parameter.

\item \emph{Palcom Assemblies} - PalCom Assemblies are rules explaining how PalCom commands shall be routed in a PalCom network. An assembly connects a service out-commands with another service in-commands and vice-versa. This is what makes it possible for different services to communicate and interact with each other. An assembly connecting services with each other could have the following rule:  “when service s1 sends command cmdOut, send it to cmdIn on service s2". Assemblies are meant to make the interconnection between different services easy to implement.

\item \emph{Palcom device} - A Palcom device is a container that runs one or more service. Had a unique id called \emph{deviceId}.
\end{itemize}

Each service has a mechanism which makes it possible to discover other services and to advertise its existence. This mechanism is called the \emph{PalCom Runtime Component}. Different services can run on an application called theThing. theThing is a type of Palcom device. It has a JVM for running services and an interpreter for assemblies. It also handles all the network configuration and connections.

@@

% TODO: Skapa en bättre och tydligare rubrik.

In this section, different commercial activity tracking systems on the market are introduced. There exists a wide range of different fitness tracking products on the market. Polar, Fitbit, Garmin, Samsung, to name a few, are all companies which have released one or several fitness devices. Many of the products has more or less the same functionality. Basic features they have in common are presented in section \ref{sec:Wearable}. The data gathered by the fitness devices needs to be stored somewhere. Three platforms that aggregates different health and fitness data are therefore also presented in this section.

One of the goals with this thesis is to evaluate in what ways the Sony Lifelog API can be used in the health care. The descriptions provided in this chapter will therefore be used as a foundation for the analysis of the Lifelog API in section\ref{sec:utLifelog}.

\section{Wearable technology}
Wearable health and fitness devices are becoming more and more popular. Smart wristbands are the most popular device in this category to use. It is predicted that the smart watches will become favored over the wristbands in 2015\cite{gartner}. The difference between a smart wristband and a smartwatch is a bit of a grey area. A smart wristband and a smartwatch have more or less the same functionality. Smart watches in general tend to have more functionality than a wristband, but this is not always the case.

There exists a wide range of types of wearable health and fitness devices. A new technology in this area is "Smart clothing", where different garments can track data such as movements and heart rate\cite{callaway2015quantified}. Progress has also been made in the fields of biosensors. Scientists at the University of San Diego have developed a tattoo-like sensor which can measure the lactate levels in the blood by converting sweat into electricity\cite{tattoo-device}.

\subsection{Smart wristbands}
\label{sec:Wearable}
In this section an overview of the different features that exists in the smart wristbands on the market is given. The focus is on wristbands but the descriptions also holds for smart watches.

Most or all of the smart wristbands on the market needs to be connected to a smartphone. The synchronization to the smartphone is performed using Bluetooth. Many of the wristband providers also synchronizes the data to their own web servers, where the data sometimes is accessible through an API. Some of these API:s are free to use and for some API:s a licence is needed.

Movement tracking and sleep tracking are two common features in the smart wristbands that exists on the market. According to Quantified Wellness, an article published by RGA Reinsurance Company\cite{callaway2015quantified}, most of the data are collected through an accelerometer in the device. Many of the smart wristbands also takes advantages of the GPS in the smartphone they are connected to in order to increase the accuracy of the measurements.

Other features that exists in wristbands on the market are the ability to track a user's heart rate, body temperature, blood oxygen level, respiration rate and galvanic skin response. The ways these sensors are used in different wristbands differ. In some wristbands, a sensor must be activated by the user in order to make a measurement and in some other wristbands data are collected without any interaction with the user at all.

>>>>

\title{RSAT}
\subtitle{A system for activity tracking of patients with kidney failure.}
%\company{The Corporation AB LTD Inc}
\supervisors{John Deer, \href{mailto:jdeer@company.com}{\texttt{jdeer@company.com}}}{Don Jeer, \href{mailto:djeer@xy.lth.se}{\texttt{djeer@xy.lth.se}}}
\examiner{Jane Doe, \href{mailto:jane.doe@cs.lth.se}{\texttt{jane.doe@cs.lth.se}}}

@@

\theabstract{
Many people suffering from kidney failure are in need of a kidney transplant. A problem in the health care is that patients cannot undergo surgery if they have too much belly fat. Regular exercises and physical activities are therefore something crucial for this group of people. A project group at the Skåne University Hospital has recently been established to help this group of people to loose weight. A question they asked themselves was whether it was possible to use any activity tracking devices in this project.  The purpose of this thesis is to design and evaluate a system that can be used to track the physical activities of patients with kidney failure. The system will be built with help of a Sony SmartBand, a wristband that collects and analyzes data about users daily activities. 
}

\keywords{kidney failure, activity tracking, loose weight, Sony Lifelog}

@@

\chapter[Introduction]{Introduction}

This thesis presents a system that aims to help over-weight patients with kidney failure to loose weight. The system is called the Region Skåne Activity tracking, or the RSAT. itACiH is a research project that investigates the possibilities of advanced health care at home of the patients. As a part of this project a planning-system was developed that is currently being used at the hospitals in Skåne. The planning system is built as a web application and can remotely fetch data from sensors that are installed at the home of a patient. RSAT is thought to be integrated with the planning system.

Many people suffering from kidney failure are in need of a kidney transplant. A problem in the health care is that patients cannot undergo surgery if they have too much belly fat. It is very hard for this group of people to loose weight, as further described in section \ref{sec:kiney}. Regular exercises and physical activities are therefore something crucial for this group of people.

A project group at the Skåne University Hospital (called the RSAT-group in this thesis) has recently been established to help this group of people to loose weight. A question they asked themselves was whether it was possible to use any activity tracking devices in this project.

Sony has developed wearable devices that can track a user’s daily activities. The data collected by these devices are saved online in their database and accessible through an API called the Lifelog API. Sony wants to investigate what different use cases that exists for their wearable devices and has therefore chosen to sponsor this project with smart wristbands for the patients to wear.

@@

The following questions will be used to fulfill the purpose:

\begin{quote}
\centering
\emph{How well suited is the Sony Lifelog platform to be used in the heath care from a technical standpoint?}
\end{quote}

\begin{quote}
\centering
\emph{How well suited is RSAT, the system to be developed, to be used in the health care?}
\end{quote}
%Frågeställningarna behöver vi tyvärr fortfarande fila på. vettitusan hur vi ska göra här...

@@

\end{comment}
\section{Method}
Before the development of RSAT could start, an interview with the RSAT-group was held. Their wishes was used as input when the requirements for the system was created. 
Thereafter a proof-of-concept system was developed to verify that it was possible to use the Lifelog API.

The approach for the development was to use iterations. Each iteration resulted in a demo for the RSAT-group. During the demo feedback was given from the RSAT-group. The experience from the development of the system creates the foundation for the evaluation of both the Lifelog platform and RSAT.

Each iteration followed these steps:

\begin{enumerate}
\item Identify the requirements for the current prototype.
\item Implement functionality in the system such that the system fulfills the requirements.
\item Test the developed prototype and publish it such that others can test our software.
\item Receive feedback from testers and use it as input to the next prototype. 
\item Receive new requirements on the system and if the requirements are feasible, add them to a later iteration. 
\end{enumerate}

@@

The first part of this thesis gives an overview of the landscape that RSAT acts in. First a medical background is given, then a technical background describing the techniques used in RSAT. After that related work to the thesis is described. First describing  different warble technologies and platforms for aggregation of activity data and  systems similar to RSAT.

The second part of this thesis contains a description of the development of RSAT. From the initial thoughts in the RSAT group to the last prototype. Telling about all decisions and struggles.

@@

@@

Kidneys have the responsibility of filtering waste products from the blood in the body\cite{NKDEPBASICS}. When the kidney function is too low to filter the blood, dialysis and in the end a kidney transplant, is needed.

In this chapter an introduction to chronic kidney disease is given, beginning with a short introduction to chronic conditions in general. Chapter 2.2 discusses chronic kidney disease especially.

@@

The costs in the health care are ever increasing. A huge part of the health care and the economic costs associated with it addresses patients with chronic conditions. More than 60\% of all deaths in the world is due to this types of complications. In Sweden 80\% of all health care costs go to treating these conditions\cite{SmartCare}. A study shows that three types of lifestyle behaviors  leads to four chronic diseases. These diseases accounts for about 50\% of all the deaths worldwide, see figure \ref{fig:chronic-fig}\cite{callaway2015quantified}.

@@

Chronic kidney disease (CKD) is a gradual loss of kidney function over time\cite{KIDNEYFOUNDATION}. There are no specific symptoms of the disease but illness, tiredness and a reduced appetite are all common symptoms. As the kidney function is worsening, waste products build to high levels in the blood and may cause complications such as high blood pressure,  muscle and bone mass loss. The risk of getting other diseases such as cardiovascular disease, pericarditis and anemia is also growing.  When the waste products in a patient’s body becomes so high that the patient starts to get sick from it, the patient requires dialysis\cite{KIDNEYFOUNDATION}.%todo: kolla över mening

@@

Despite that CKD is considered to be an irreversible condition, the progress of the disease can be slowed down. The National Kidney Disease Education Program (NKDEP) lists three strategies that can be used to slow down the progress\cite{NKDEPPROG}. One of those strategies is lifestyle interventions and addresses different health-promoting behaviour. One of the health-promoting behaviours they discuss is physical activity. NKDEP states that people suffering from CKD have the same recommendations for physical activity as the general population.

Eventually, a patient with CKD must undergo a kidney transplant. The patient must be in good physical shape, as the risk of infections and other diseases otherwise are too high. It is also essential that the patient does not have too much belly fat. Not only does the fat induce a greater risk for the patient's life, it also makes it harder for surgeons to practically perform the transplant. %Vad kostar det?!
%http://www.kidney.ca/document.doc?id=4083
%http://health.costhelper.com/kidney-cancer.html

@@

This chapter describes the techniques used to develop RSAT. First the Sony Lifelog and techniques used by Lifelog are described. The second part describes PalCom which is the platform used by the itACiH and RSAT.

@@

The Sony Lifelog application is an Android application which is built around the idea of logging a users lifestyle and daily activities\cite{LifeLogDescr}. The application collects various data about its users through different sensors in a smartphone and SmartWear-devices (Sony SmartWatch or a Sony Smartband) and displays it to the user. Example of data it logs \emph{?!?!are/is?!?} information about what applications in the smartphone that have been used, what locations that have been visited and information about different physical activities performed during the day.

@@

The Sony SmartBand SWR10 is the chosen wristband for this project. The Sony SWR10 has a movement sensor that recognizes different movements a user makes\cite{SWR10}. This data is analyzed by Sony and categorized into different activates, which are uploaded to a central repository that stores the data. The SmartBand must be connected to an Android device with Android version 4.4 or higher to be able to collect data.

\subsection{Sony Lifelog API} 
All the data collected in the Sony Lifelog application are automatically uploaded to a central repository that stores the data. This data is globally accessible through a REST-based web-service API\cite{LifeLogDescr}.

The API has three different endpoints; a profile endpoint, an activities endpoint and a location endpoint. The profile endpoint contains basic information about a user's name, age, height and weight and is entered by the user into the Lifelog application. The activities endpoint contains the different data Lifelog has logged about performed activities throughout the day such as physical activity. All activities have a type and some also has a subtype.

The location endpoint provides a list of coordinates specifying what locations a user has been visiting during a certain time period. Activities and locations in the API are associated with times. A response from the API always includes the time stamp of the earliest and latest included entry. The format of the timestamp follows the ISO 8601 standard.

Third parties can access the API to retrieve collected data about a user. Every request made to the API must be authorized and https must be the chosen communication protocol. The lifelog Web API uses OAuth 2.0 as its authorization mechanism. To connect to the API, a user must login and give explicit permission for the application or the service to access to the data.

The response to a valid request made to the API contains a result-section. This result-section is a set of data represented as JSON objects. If the requested data is to large to fit in the size of one response message, the data is split up into several messages and a link-field is inserted into the response. This link-field contains a link pointing to the location from where the rest of the data can be fetched.

@@

The Sony Lifelog API is a REST API. REST is an architectural style for designing API:s and are heavily used in modern web services. The concept is to represent resources as the analogy for the data endpoints and to have a small set of actions interacting with these resources. The REST architecture is based on the following principles\cite{pautasso2014restful}:

\begin{itemize}
    \item \emph{Addressabilty} All resources must be addressable with a unique identification. On the web each resource uses a unique URI such that it becomes globally addressable.
    \item \emph{Uniform Interface} The interface to the resources should according to REST be uniformed and simple. On the web the HTTP protocol is used and HTTP methods are used to interact with resources. The GET method is used to read a resource, the POST method is used to add a resource, the PUT method is used to to modify a resource and the DELETE method is used to delete a resource.
    \item \emph{Stateless Interaction} This means that there is no session with multiple interactions. Instead every interaction is self-contained. No state is shared between a service and client after an interaction.
    \item \emph{Self-Describing Messages} The requests and responses should contain both the data and the metadata about the format of the data. XML or JSON is usually used to represent data on the web. The Sony Lifelog API uses JSON to represent its data.
\item \emph{Hypermedia} Resources that contains a link to other resources uses hypermedia. If the data (the resource) to be sent in a response is too large to fit in a single message, the data can be organized in several messages, where each message contains a link (hypermedia) to the next part of the resource. Then, only the first part of the resource is sent. The receiver can later fetch the rest of the data with help from the hypermedia link in the message.

@@

OAuth 2.0 is an authorization protocol developed by the Internet Engineering Task Force (IETF). It describes an open standard for authorization and enables third party applications to obtain limited access to a web-service on the behalf of a resource owner\cite{oauth2spec}. To describe the protocol, several key actors need to be described, these actors are the client, the resource owner, the authorization server and the resource server. IETF gives the following definitions to the roles:

@@

The following is a general example of the protocol flow when a client is accessing protected data belonging to a user.

\begin{enumerate}
	\item A client requests authentication from a resource owner. 
	\item The client receives an authorization grant credential from the resource owner.    
	\item The client requests an access token from the authorization server. The server verifies the authorization grant credential and returns an access token. 
	\item The client receives an access token, which he/she can use to access the protected data in the resource server. 
\end{enumerate}

The Authorization Grant is a credential representing the resource owner's authorization. The IETF's OAuth 2.0 specification defines four different grant types; authorization code, implicit, resource owner password credentials and client credentials. The Sony Lifelog API uses the authorization code as its grant type and the other types will therefore not be further discussed\cite{LifeLogAuth}. When the authorization code grant type is used, the resource owner is redirected to an authentication server when the client requests authentication. When the resource owner has granted the request, the resource owner is redirected back to the client together with the authorization code, which is appended in the URL. The authentication code is short lived and single-use, meaning that if the code is used more than once, the access tokens related to the code will be revoked.

The access token is the only credential needed to access a user's protected resource and can be seen as an abstraction layer, since other credentials such as user-name and password must not be provided. An access token is valid until it expires or is revoked.

A refresh token can be issued together with an access token and, as the name suggests, is used to refresh expired or revoked access tokens. To refresh an access token, the refresh token is sent to the authorization server, which issues a new access token. A refresh token is meant to be valid for a longer time than an access token but that depends on the server implementing the OAuth-specification. A refresh token could in theory expire before an access token.

\section{PalCom}
The planning system in the itACiH project use PalCom as their platform. PalCom is a platform that makes it possible to build software systems, which consists of small sub-systems\cite{ist-PalCom}. These sub-systems are called \emph{PalCom Services}. Different services are composed together with the help of assemblies. Some of the most common and important components in the PalCom platform are:

\begin{itemize}
\item \emph{PalCom service} - A PalCom service is a type of self-contained system. The service contains logic for both computations and interactions with the outside world. The implementation of services is made in Java and can therefore leverage the whole Java ecosystem. Every version service has unique id called \emph{serviceId} and every instance of a service has an unique id called \emph{instance}.

\item \emph{PalCom command} - A PalCom command is what a PalCom Service uses to communicate with other services. A service defines a number of commands, either \emph{in-commands} or \emph{out-commands}. An in-command defines the format of a command sent to the service and an out command defines the format of a command sent from the service. A command consists of zero or more parameters. Each parameter has a name and a data type. The data type defines what type of data is sent in the parameter.

\item \emph{PalCom assemblies} - PalCom Assemblies are rules explaining how PalCom commands shall be routed in a PalCom network. An assembly connects a service out-commands with another service in-commands and vice-versa. This is what makes it possible for different services to communicate with each other. An assembly connecting services with each other could have the following rule:  “when service s1 sends command cmdOut, send it to cmdIn on service s2". Assemblies are meant to make the interconnection between different services easy to implement.

\item \emph{PalCom device} - A PalCom device is a container that runs one or more services and has a unique id called \emph{deviceId}.

\item \emph{PalCom Runtime Component} - Each service has a mechanism, which makes it possible to discover other services and to advertise its existence. This mechanism is called the PalCom Runtime Component. Different services can run on an application called theThing. theThing is a type of PalCom device. It has a JVM for running services and an interpreter for assemblies. It also handles all the network configuration and connections.

@@

In this section, different commercial activity tracking systems on the market are introduced. There exists a wide range of different fitness tracking products on the market. Polar, Fitbit, Garmin, Samsung, to name a few, are all companies, which have released one or several fitness devices. Many of the products have more or less the same functionality. Basic features they have in common are presented in section \ref{sec:Wearable}. The data gathered by the fitness devices needs to be stored somewhere. Three platforms that aggregates different health and fitness data are therefore also presented in this section.

\section{Wearable technology}

Wearable health and fitness devices are becoming more and more popular. Smart wristbands is the most popular devices  in this category, though it is predicted that the smart watches will become favored over the wristbands in 2015\cite{gartner}. The difference between a smart wristband and a smartwatch is a bit of a grey area. A smart wristband and a smartwatch have more or less the same functionality. Smart watches in general tend to have more functionality than a wristband, but this is not always the case.

There exists a wide range of wearable health and fitness devices. A new technology in this area is "smart clothing", where different garments can track data such as movements and heart rate\cite{callaway2015quantified}. Progress has also been made in the fields of biosensors. Scientists at the University of San Diego have developed a tattoo-like sensor, which can measure the lactate levels in the blood by converting sweat into electricity\cite{tattoo-device}.

\subsection{Smart wristbands - basic features}
\label{sec:Wearable}
The focus in this section is on wristbands but the descriptions also holds for smart watches.

Movement tracking and sleep tracking are two common features in the smart wristbands that exist on the market. According to Quantified Wellness, an article published by RGA Reinsurance Company\cite{callaway2015quantified}, most of the data are collected through an accelerometer in the device. Many of the smart wristbands also takes advantages of the GPS in the smartphone they are connected to in order to increase the accuracy of the measurements.

Other features that exist in wristbands on the market are the ability to track a user's heart rate, body temperature, blood oxygen level, respiration rate and galvanic skin response. The ways these sensors are used in different wristbands differ. In some wristbands, the user must activate a sensor in order to make a measurement and in some other wristbands data are collected without any interaction with the user at all.

The most common type of heart rate sensor is optical and uses LED-lights and an optical sensor to detect the amounts of blood flowing through the wrist in order to calculate a user’s pulse. The accuracy of different optical heart rate sensors varies a lot. In a test performed by CNET, four fitness trackers with an optical heart rate sensor were used to measure the heart rate. The results were then compared to the pulse measured by an EKG machine. The variance in the results were big.  At a pulse rate of 80 – 90 bpm, all of the devices was able to measure the heartbeat and was at the most 10\% off from the real heart rate. At a heart rate of 160 – 170 bpm, one of the trackers could not measure any data at all and two of the trackers were more than 50\% off\cite{CNET}. One of the devices used in the test, a Samsung Galaxy S5 smartphone, proved to be very accurate. It was at most 3\% off compared to the EKG-result.
